---
title: "EDS 222 Homework 2"
author: "Gabe Ingman, with help from Ray Hunter"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load all the packages needed here
library(tidyverse)
library(here)
library(readr)
library(gt)
library(tufte)
library(cowplot)
library(ggplot2)

# Set your filepaths here! Or, set this up as an .Rproj if you'd like.

#It's an R Project
```

# Question 1: Probability density functions in `R`

`R` has many built-in functions that let you describe, analyze, and sample from common probability density functions. For example, if you type `?stats::Normal` in your console, you'll see documentation on all the functions relevant to the normal distribution.[^1] These functions include:

[^1]: Recall that the normal distribution is a family of distributions that are symmetric and do not have long tails. They each have different means $\mu$ and standard deviations $\sigma$.

-   `dnorm(x, mean = 0, sd = 1)`, which returns the *density* of the normal distribution evaluated at whatever values you pass in via `x`. You can think of the output of `dnorm()` as the *height* of the normal pdf at the value `x`. Note that this function defaults to a normal distribution with $\mu = 0$ and $\sigma = 1$, but you can of course change that.

-   `pnorm(q, mean = 0, sd = 1)`, which returns the *cumulative probability* of the normal distribution evaluated at whatever values you pass in via `q`. You can think of the output of `pnorm()` as the *area* under the pdf to the left of the value `q`. Again, note the default distribution parameters $\mu$ and $\sigma$.

## Question 1.1

```{R, echo = T}
x = seq(-4, 4, 0.01)
```

Use `dnorm()` to compute the density of the normal pdf for all values in the `x` vector generated above, using $\mu = 0$ and $\sigma = 1$. Use `geom_polygon()`, `geom_line()`, or `geom_point()` (take your pick) to plot this pdf over the support given in `x`.

```{R, echo = TRUE, message=FALSE, warning=FALSE}

#getting density of the normal pdf for all values of 'x'
x_dnorm <- dnorm(x = x, 
                 mean = 0, 
                 sd = 1)
#x and y density values dataframe
df_x_dorm <- data.frame(x, x_dnorm)

#graphing the dataframe
ggplot(df_x_dorm, aes(x = x, y = x_dnorm)) +
  geom_line(color = "blue")

```

## Question 1.2

Use the densities you generated in 1.1 to calculate the probability that a random variable distributed normally with mean 0 and standard deviation 1 falls between -2 and 2.[^2]

[^2]: Hint: Remember that $$ Pr(A\leq x \leq B) = \int_A^B f(x)dx $$ where the integral is a fancy way to tell you to sum up $f(x)$ over values of $x$ from $A$ to $B$.

**The probability that a random variable distributed normally falls between -2 and 2 is 95.%.**

```{r}

#finding probability that it falls within 2
prob_2 <- pnorm(2, mean = 0, sd = 1, lower.tail = TRUE)
prob_2

#finding the probability that it falls under negagive 2
prob_neg2 <- pnorm(-2, mean = 0, sd = 1, lower.tail = TRUE)
prob_neg2

# subtracting the probability that it falls under negative two by two to ensure to find the probability of a random variable that falls between the two
prob_neg2_2 <- prob_2 - prob_neg2
prob_neg2_2

```

## Question 1.3

Suppose $\sigma=2$ instead. Qualitatively, how would your answer to Question 1.2 change? Why?

**If the standard deviation was 2 instead of 1, the chance that a random variable normally distributed would be lower. Changing the standard deviation would change the shape of the graph and the distribution would be more flat,more distributed bell curve.**

## Question 1.4

An analogous set of functions computes densities and probabilities for the **log normal** distribution. These functions are `dlnorm()` and `plnorm()` and operate as above for the normal distribution functions.

Use `plnorm()` under default parameters to compute the probability that a random variable distributed log normal takes on a value above 2. Use `pnorm()` to compute the corresponding probability for the normal distribution under default parameters. Why are these values so different?

```{r}

plnorm_above2 <- 1 - plnorm(2, mean = 0, sd = 1)
plnorm_above2


pnorm_above2 <- 1 - pnorm(2, mean = 0, sd =1)
pnorm_above2

```

**The different distributions create different shapes, that create different skews and tails. There is a difference probability, even with a similar data set, in trying to find a random variable, distributed 'log normal', that is >2, than the probability that a random variable, distributed normally, is >2. I am not sure if I explained that correctly but that is my understanding at this time.**

# Question 2: Climate summary statistics

In the following questions, you'll be working with climate data from Colombia. These data were obtained from the [ERA5 database](https://www.ecmwf.int/en/forecasts/datasets/reanalysis-datasets/era5), a product made available by the European Centre for Medium-Range Weather Forecast. The high-resolution hourly gridded data were aggregated to the municipality by month level -- that is, each observation in these data report a monthly average temperature value and a monthly cumulative precipitation value for one of the 1,123 municipalities across the country.[^3]

[^3]: Note: The computational techniques we use to go from raw, spatial, gridded data to a tabular dataset at an administrative level are really valuable for environmental data science. Between Ruth and I, we're hoping to cover some of these topics later in the quarter!

These data -- stored in `colombia_climate.csv` -- cover all municipalities for the period 1996 to 2015. Climate scientists tend to describe the "climate" of a location as the probability density function of a large set of climate variables over about a 30 year period. We only have 20 years, but we will consider our sample as randomly drawn temperature and precipitation realizations from the "climate" p.d.f. over this period. We are aiming to draw conclusions about the Colombian climate using this sample of temperature and precipitation observations.

## Question 2.1

Read these data into `R` using the `read.csv()` function.[^4]

[^4]: See the README.rtf file for details on the variables in `colombia_climate.csv`.

For each of the temperature and rainfall variables, create a histogram that shows the distribution of the variable across the entire sample. For each variable, answer the following questions:

-   Is the distribution symmetric or skewed?
-   Is there a long tail (or two), or does this distribution look approximately normally distributed?
-   Is the distribution unimodal, bimodal, or multimodal?

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\--

**Precipitation Variable:**

-   **The distribution is asymmetrical, and skewed right.**

-   **As rainfall increases, there is a longer tail.**

-   **The distribution is unimodal.**

**Temperature Variable:**

-   **The distribution is asymmetrical, and skewed right.**

-   **The temperature data does not look normally distributed. Since values above 24 C are considered part of the tail to the right, it does not seem normal.**

-   **The distribution is bimodal given that the values with a count of 10,000+ can be considered a second mode.**

```{r, fig.fullwidth=TRUE, fig.height=4, message=FALSE, warning=FALSE}

colombia_climate <- read.csv(here("colombia_climate.csv"))

#preciptation histogram
precip_histogram <- ggplot(data = colombia_climate, aes(x = precip)) +
  geom_histogram(fill = "cornflowerblue",
                 color = "turquoise") +
  labs(x = "rainfall (mm)",
       y = "count",
       title = "Avg Monthly Rainfall in Colombia")
precip_histogram

#temperature histogram
temp_histogram <- ggplot(data = colombia_climate, aes(x = temperature)) + 
  geom_histogram(fill = "turquoise3",
                 color = "blue") +
  labs(x= "temperature (C)",
       y = "count",
       title = "Avg Monthly Temperature in Colombia")
temp_histogram

mean(colombia_climate$temperature)
median(colombia_climate$temperature)

mean(colombia_climate$precip)
median(colombia_climate$precip)
```

## Question 2.2

Given your answers to 2.1 above, do you expect the mean of temperature to differ from the median? Is it likely to be about the same, smaller, or larger? What about precipitation?

\-\-\-\-\-\-\-\-\-\-\-\-\-\--


**Precipitation: It's reasonable to assume that the mean would be higher than the median. There are more values on the lower end of the dataset, meaning the middle value / median might be lower, while the average value / mean might be higher.**

**Temperature: This is my best guess: the median might be a larger value than the mean because of the bimodal distribution, that skews the median higher.**

## Question 2.3

Anthropogenic climate change is expected to raise temperatures across Colombia, increase total precipitation, and increase variability in precipitation. Compute the mean, the median, and the standard deviation of each climate variable in:

-   All years before and including 2005
-   All years after 2005

Put your summary statistics into a table (or two tables, whatever is easiest). Are the changes you see between the pre-2005 and post-2005 periods consistent with climate change? Explain why.

**Precipitation variability increased, as shown by the standard deviation increasing almost 40 mm from pre-2005 to post-2005. This lines up with what I know about the effects of climate change on countries near the equator: more intense and unpredictable weather. However, the average temperature stayed about the same pre-and-post 2005, which does not line up with climate change. All of the hottest years on record, in terms of temperature, have happened post-2012.**

```{r, fig.fullwidth=TRUE, fig.height=4, message=FALSE, warning=FALSE}

#creating separate dataframes for data pre-2006 and post 2005 precipitation
colombia_climate_pre2005 <- filter(colombia_climate, year <= 2005)
colombia_climate_post2005 <- filter(colombia_climate, year > 2005)

# summary for pre 2005 precipitation
mean_pre2005_precip <- mean(colombia_climate_pre2005$precip)
med_pre2005_precip <- median(colombia_climate_pre2005$precip)
sd_pre2005_precip <- sd(colombia_climate_pre2005$precip)

# summary for pre 2005 temperature
mean_pre2005_temp <- mean(colombia_climate_pre2005$temperature)
med_pre2005_temp <- median(colombia_climate_pre2005$temperature)
sd_pre2005_temp <- sd(colombia_climate_pre2005$temperature)

# summary for post 2005 precipitation
mean_post2005_precip <- mean(colombia_climate_post2005$precip)
med_post2005_precip <- median(colombia_climate_post2005$precip)
sd_post2005_precip <- sd(colombia_climate_post2005$precip)

# summary for pre 2005 temperature
mean_post2005_temp <- mean(colombia_climate_post2005$temperature)
med_post2005_temp <- median(colombia_climate_post2005$temperature)
sd_post2005_temp <- sd(colombia_climate_post2005$temperature)

#concatinating the values for each year and variable
pre2005_precip <- c(mean_pre2005_precip, med_pre2005_precip, sd_pre2005_precip)
pre2005_temp <- c(mean_pre2005_temp, med_pre2005_temp, sd_pre2005_temp)
post2005_precip <- c(mean_post2005_precip, med_post2005_precip, sd_post2005_precip)
post2005_temp <- c(mean_post2005_temp, med_post2005_temp, sd_post2005_temp)

#creating new names for the values
values <- (c("mean", "median", "standard deviation"))

#dataframe for the precipitation variables
table_rainfall <- data.frame(values, pre2005_precip, post2005_precip) %>% 
  gt
table_rainfall

#combining in a dataframe and then producing a table for the precipitation varibale
table_temperature <- data.frame(values, pre2005_temp, post2005_temp) %>% 
  gt
table_temperature

```

## Question 2.4

The histograms and summary statistics should make you concerned that these data are not normally distributed. As we will show later in the course, it's often very helpful to have normally distributed data before we do things like linear regressions or hypothesis testing. Here, let's use a Q-Q plot to assess the normality of our sample data.

-   Use `geom_qq()` and `geom_qq_line()` in `ggplot2` to make a Q-Q plot for each variable.[^5]

-   What do you conclude about the normality of these two variables?

[^5]: `geom_qq_line()` lets you draw a line indicating where the sample quantiles would lie if the data were normally distributed.

```{r, message=FALSE, warning=FALSE}

#Graphing the temperature with a best-fit trendline
qq_temp <- ggplot(data = colombia_climate, aes(sample = temperature)) +
  geom_qq() +
  geom_qq_line()+
  xlab("temperature")
qq_temp

#Graphing the temperature with a best-fit trendline
#Note- I forgot how to add a title to a QQPlot so I added a label to the x-axis. Consider the x-axis label as the title for the graph.
qq_precip <- ggplot(data = colombia_climate, aes(sample = precip)) +
  geom_qq() +
  geom_qq_line()+
  xlab("precipitation")
qq_precip
```

**Temperature: not normal. The trend line is not a good fit.**

**Precipitation: not normal. The trend line is not a good fit**

## Question 2.5

When our sample observations are not normally distributed, we often rely on nonlinear transformations[^6] to reshape our data. If we compute a nonlinear transformation on our underlying data and they then look closer to normal, we can use this transformed version of our variable in later statistical analysis.

[^6]: Any mathematical operation that is a nonlinear function of the underlying variable can be considered a "nonlinear transformation". For example, $x^2$ and $log(x)$ are both nonlinear transformations.

Because we tend to see a lot of variables in the world that follow the log normal distribution, a very common nonlinear transformation is the natural logarithm. Transform the precipitation data by taking the natural logarithm. Then remake your Q-Q plot -- does your variable (defined as `log(precip)`) now look closer to normally distributed? What can you learn about where the data diverge from the normal distribution?

```{r, message=FALSE, warning=FALSE}

log_data <- colombia_climate |> 
  mutate(log_precip = log(precip)) |> 
  mutate(log_temp = log(temperature))

qq_precip_log <- ggplot(data = log_data, 
                    aes(sample = log_precip)) +
  geom_qq() +
  geom_qq_line()+
  xlab("precipitation transformed by log")

qq_precip_log
```

**The precipitation, when transformed by the log function, appears much closer to a normal distribution. This means the data is distributed normally only when it's **

